---
title: "Individual Assignment 5"
author: "Stella Freedberg"
date: "05-10-2026"
format: pdf
---

# Set Up

## Packages 

```{r}
#| label: packages
#| message: false

# general use and cleaning 
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading .xlsx files 
library(readxl)

# visulaize missing data
library(naniar)

# arrange plots 
library(patchwork)

```

## Data 

```{r}
#| label: data
#| message: false

# taxonomic information
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
  ) 

# site information
sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quality 
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c(" ", "n/a", "N/A", "over", "173+")
)

```

# Cleaning Data 

```{r}
#| label: NCOS-sites

# creating new object from sites 
ncos_sites <- sites |> 
  # filter to only include NCOS quaerterly sampling sites 
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

```{r}
#| label: cleaning-taxon

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # convert taxon names to snakecase
  mutate(verbatim_name = to_any_case(verbatim_name, 
                                     case = "snake"))
```

```{r}
#| label: water-quality-cleaning

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 
```

```{r}
#| label: cleaning-aq-ins
#| message: false

aq_ins_clean <- aq_ins |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  rename (date = date_on_vial) |> 
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         nematode,
         diptera_ceratopogonidae,
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  filter(sample_type %in% c("FB250", "FB 250")) |> 
  pivot_longer(
    # columns to make lomger 
    cols = ostracod:annelida_polychaete,
    # new column for the old column names
    names_to = "taxon",
    # new column for the values in each of the old columns
    values_to = "abundance"
    ) |> 
  group_by(date, site, taxon) |> 
  summarize(ave_lit = sum(abundance, ra.rm = TRUE)/7.5) |> 
  ungroup() |> 
   # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  ))

```

```{r}
#| label: visualizing-salinity
#| warning: false
#| fig-height: 6
#| fig-width: 8

# base layer
salinity <- ggplot(data = aq_ins_clean,
       # x-axis: site
       # y-axis: mean salinity
       mapping = aes(x = site_full,
                     y = med_sal)) +
  # first layer: salinity measurements at each survey
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # second layer: points to represent means
  stat_summary(fun = median,
               color = "red",
               size = 1) +
  # relabelling x- and y-axes
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site salinity") +
  # different theme
  theme_bw()

```

```{r}
#| label: copepod-cleaning
#| warning: false

# creating new object from clean aquatic inverts 
copepods <- aq_ins_clean |> 
  # filter to only include copepods
  filter(taxon == "copepod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1)) 
```

```{r}
#| label: visualizing-copepods
#| warning: false
#| fig-height: 6
#| fig-width: 10

# base layer
copepods <- ggplot(data = copepods,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "blue") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(15)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Copepods") +
  # different theme
  theme_bw()

```

# 1. Visulaizing differences across sites in major taxon abundance 

## a)

- x-axis: site 
- y-axis: log + 1 transformed average abundance/L
- geom to represent average abundance/L: geom_jitter()
- geom to represent medians: stat_summary()

## b) 

```{r}
#| label: wrangle-aq-ins-clean
#| message: false
#| warning: false

# creating new object from clean aquatic inverts 
ostracod <- aq_ins_clean |> 
  # filter to only include ostracod
  filter(taxon == "ostracod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1)) 

ostracod |> 
  slice_sample(n = 10)

```

## c)

```{r}
#| label: visualizing-ostracod
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 10

# base layer
ostracod_plot <- ggplot(data = ostracod,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "pink") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(15)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Ostracods") +
  # different theme
  theme_bw()

# displaying object
ostracod_plot

```

## d) 

```{r}
#| label: create-multipanel-figure
#| warning: false
#| fig-height: 10
#| fig-width: 15

# create multipanel figure
salinity / (copepods + ostracod_plot)

```

## e) 

Copepods are more abundant than ostracods at these sites: Devreux Creek, North of Venoco Bridge, Phelps Road, and South of Venoco Bridge. 

It does not appear that ostracod abundance relates to site-level differences in salinity. At South of Venoco Bridge salinity is high while abundance is low however at North of Venoco Bridge both salinity and abundance are high. 

# 2. Visualizing total abundance as a function of salinity 

## a)

- x-axis: med_sal
- y-axis: log_ave_lit
- color: taxon
- geometry: geom_jitter()
- facets: facet_wrap( ~ taxon, scales = "free")

## b) 

```{r}
#| label: wrangle-aq-ins-salinity
#| message: false
#| warning: false

# creating new object from clean aquatic inverts
aq_ins_salinity <- aq_ins_clean |>
  # create log + 1 transformed average abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1),
         # make taxon names title case
         taxon = str_to_title(taxon)) |>
  # order taxon levels by maximum abundance
  mutate(taxon = fct_reorder(taxon, ave_lit, .fun = max, .desc = TRUE)) |>
  # keep columns needed for figure
  select(taxon, ave_lit, log_ave_lit, med_sal, site_full)

# display 10 random rows
aq_ins_salinity |>
  slice_sample(n = 10)

```

## c)

```{r}
#| label: code-your-figure

salinity_taxa <- ggplot(data = aq_ins_salinity,
                        mapping = aes(x = med_sal,
                                      y = log_ave_lit,
                                      color = taxon)) +
  geom_point() +
  # separate panel for each taxon
  facet_wrap(~ taxon, scales = "free") +
  # custom colors 
  scale_color_brewer(palette = "Dark2") +
  # labels
  labs(x = "Median salinity (ppt)",
       y = "log + 1 transformed average abundance/L") +
  # remove legend
  theme_bw() +
  theme(legend.position = "none")

# display figure
salinity_taxa
```

